---
title: "Cultural Matching Analysis"
author: "Omar Lizardo et al."
date: today
bibliography: manuscript_citations.bib
format: 
  html:
    toc: true
    keep-tex: true
    theme: cosmo
  pdf:
    toc: true
    keep-tex: true
execute:
  echo: false
  warning: false
  cache: true
---
## Overview
This analysis file examines the role of cultural matching and network opacity in predicting the persistence of social ties over time. We pool data across adjacent waves from the NetSense study to construct a person-period dataset, restricting the sample to non-kin, on-campus student ties.
We estimate the probability of tie persistence as a function of cultural matching (closed-form domains and open-ended activities) and network opacity (the number of "Don't Know" responses regarding an alter's preferences) using mixed-effects logistic regression (`lme4::glmer`).
```{r}
#| label: setup
#| include: false
library(tidyverse)
library(lme4)
library(modelsummary)
library(here)
library(broom.mixed)
ego_race <- readRDS(here("data", "processed", "ego_race.rds")) %>% mutate(sender = as.character(sender))
alter_race <- readRDS(here("data", "processed", "alter_race.rds")) %>% mutate(receiver = as.character(receiver))
df_period <- readRDS(here("data", "processed", "adjacent_waves.rds")) %>%
  filter(kin == 0, campustie == 1, !is.na(alterid)) %>%
  left_join(ego_race, by = c("egoid" = "sender")) %>%
  left_join(alter_race, by = c("alterid" = "receiver")) %>%
  mutate(
    period = factor(period),
    persisted = as.numeric(tie_persist > 0),
    female_factor = factor(female, levels = c(0, 1), labels = c("Men", "Women")),
    alterfemale_factor = factor(alterfemale, levels = c(0, 1), labels = c("Men", "Women")),
    close_factor = factor(close, levels = c(3, 2, 1), labels = c("Not Close", "Somewhat Close", "Close")),
    same_dorm = as.numeric(reltype == 3 | reltype == 4),
    is_friend = as.numeric(reltype == 6),
    race_homophily = case_when(
      is.na(ego_race) | is.na(alter_race) ~ "Unknown",
      ego_race == alter_race ~ "Homophilous",
      TRUE ~ "Heterophilous"
    ),
    race_homophily = factor(race_homophily, levels = c("Heterophilous", "Homophilous", "Unknown")),
    freq_high = if_else(freq_factor %in% c("Weekly", "Daily"), 1, 0, missing = NA_real_),
    duration_sq = duration_^2
  ) %>%
  filter(!is.na(same_dorm), !is.na(freq_high))
glmer_ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
ctrl_vars <- "same_dorm + is_friend + race_homophily + freq_high + female_factor + alterfemale_factor + close_factor + duration_ + duration_sq + period"
mod_closed <- glmer(
  as.formula(paste("persisted ~ num_match_closed +", ctrl_vars, "+ (1 | egoid)")),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)
mod_open <- glmer(
  as.formula(paste("persisted ~ num_match_closed + open_match_count +", ctrl_vars, "+ (1 | egoid)")),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)
mod_opac_base <- glmer(
  as.formula(paste("persisted ~ num_match_closed + open_match_count + num_unknown +", ctrl_vars, "+ (1 | egoid)")),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)
```
## Main Effects Models
The following models assess the main effects of closed-form cultural matching, open-ended cultural matching, and network opacity on the odds of tie persistence. Model 1 shows that closed-form matches independently predict persistence. Model 2 introduces open-ended matches, demonstrating the value of incorporating both open and closed-form items. Model 3 incorporates network opacity (the number of "Don't know" responses regarding an alter's preferences). The main effects model reveals that a lack of shared cultural knowledge signals a weaker tie.
```{r}
#| label: tbl-models
#| tbl-cap: "Odds Ratios for Tie Persistence (Main Effects Models)"
modelsummary(
  list("Model 1: Closed" = mod_closed, "Model 2: + Open" = mod_open, "Model 3: + Opacity" = mod_opac_base),
  stars = TRUE,
  exponentiate = TRUE,
  title = "Odds Ratios for Tie Persistence (Main Effects Models)",
  output = here::here("Tabs", "main_models.tex"),
  coef_map = c(
    "num_match_closed" = "Closed-Form Matches",
    "open_match_count" = "Open-Ended Matches",
    "num_unknown" = "Network Opacity",
    "close_factorSomewhat Close" = "Subjective Closeness: Somewhat",
    "close_factorClose" = "Subjective Closeness: Close",
    "same_dorm" = "Roommate/Dormmate",
    "is_friend" = "Friend",
    "freq_high" = "Frequency: Weekly/Daily",
    "race_homophilyHomophilous" = "Race Homophily: Same Race",
    "duration_" = "Tie Duration (Years)"
  ),
  gof_map = c("nobs", "aic", "bic"),
  notes = "Note: All models control for ego and alter gender, non-linear tie duration, and baseline hazard variations across all 7 wave transitions (coefficients omitted for space). Random intercepts for egos are included."
)
```
### Marginal Effects
**Figure @fig-combined** visualizes the marginal effect (calculated as the change from the minimum to the maximum value) of closed-form matches, open-ended matches, and network opacity on the predicted probability of tie persistence.
```{r}
#| label: fig-combined
#| fig-cap: "Marginal Effects of Cultural Matches and Network Opacity (Min to Max)"
library(marginaleffects)

comp_closed <- comparisons(
  mod_open,
  variables = list(num_match_closed = range(df_period$num_match_closed, na.rm = TRUE)),
  newdata = datagrid(),
  re.form = NA
) %>% as_tibble() %>% mutate(Predictor = "Closed-Form Matches")

comp_open <- comparisons(
  mod_open,
  variables = list(open_match_count = range(df_period$open_match_count, na.rm = TRUE)),
  newdata = datagrid(),
  re.form = NA
) %>% as_tibble() %>% mutate(Predictor = "Open-Ended Matches")

comp_opacity <- comparisons(
  mod_opac_base,
  variables = list(num_unknown = range(df_period$num_unknown, na.rm = TRUE)),
  newdata = datagrid(),
  re.form = NA
) %>% as_tibble() %>% mutate(Predictor = "Network Opacity (Unknowns)")

comp_combined <- bind_rows(comp_closed, comp_open, comp_opacity) %>%
  mutate(Predictor = factor(Predictor, levels = rev(c("Closed-Form Matches", "Open-Ended Matches", "Network Opacity (Unknowns)")), ordered = TRUE))

p <- ggplot(comp_combined, aes(y = Predictor, x = estimate, fill = Predictor, color = Predictor)) +
  geom_col(width = 0.3, alpha = 0.8) +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.1, linewidth = 1) +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c("Closed-Form Matches" = "#1f78b4", "Open-Ended Matches" = "#33a02c", "Network Opacity (Unknowns)" = "#e31a1c")) +
  scale_color_manual(values = c("Closed-Form Matches" = "#12476b", "Open-Ended Matches" = "#1e5e1a", "Network Opacity (Unknowns)" = "#941113")) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  labs(y = NULL, x = "Average Marginal Effect\n(Prob. of Persistence)") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

ggsave(here::here("Plots", "main_effects.png"), plot = p, width = 8, height = 6)
p
```
## Subjective Closeness Interactions
To test whether the protective effects of cultural matching or the detrimental effects of network opacity depend on the strength of the tie, we estimated auxiliary interaction models between subjective closeness and each of the three cultural variables. While interaction terms in logistic regression models can be difficult to interpret directly from coefficients due to the non-linear link function, computing predicted probabilities across the range of the variables provides a clearer picture of the underlying dynamics.
```{r}
#| label: setup-closeness-interactions
#| include: false
ctrl_vars_int <- "same_dorm + is_friend + race_homophily + freq_high + female_factor * alterfemale_factor + duration_ + duration_sq + period"

mod_close_int_closed <- glmer(
  as.formula(paste("persisted ~ num_match_closed * close_factor +", ctrl_vars_int, "+ (1 | egoid)")),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)

mod_close_int_open <- glmer(
  as.formula(paste("persisted ~ open_match_count * close_factor +", ctrl_vars_int, "+ (1 | egoid)")),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)

mod_close_int_unknown <- glmer(
  as.formula(paste("persisted ~ num_unknown * close_factor +", ctrl_vars_int, "+ (1 | egoid)")),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)
```
**Figure @fig-closeness-interactions** plots the marginal effects of the three cultural variables across the three levels of subjective closeness.
```{r}
#| label: fig-closeness-interactions
#| fig-cap: "Marginal Effects of Cultural Matching and Network Opacity by Subjective Closeness"
#| fig-width: 12
#| fig-height: 5
#| warning: false
library(patchwork)

comp_int_closed <- comparisons(mod_close_int_closed, variables = list(num_match_closed = c(0, 6)), newdata = datagrid(close_factor = c("Not Close", "Somewhat Close", "Close")), re.form = NA) %>% as_tibble() %>% mutate(Cultural_Variable = "Closed-Form Matches")
comp_int_open <- comparisons(mod_close_int_open, variables = list(open_match_count = c(0, 5)), newdata = datagrid(close_factor = c("Not Close", "Somewhat Close", "Close")), re.form = NA) %>% as_tibble() %>% mutate(Cultural_Variable = "Open-Ended Matches")
comp_int_opacity <- comparisons(mod_close_int_unknown, variables = list(num_unknown = c(0, 6)), newdata = datagrid(close_factor = c("Not Close", "Somewhat Close", "Close")), re.form = NA) %>% as_tibble() %>% mutate(Cultural_Variable = "Network Opacity (Unknowns)")

comp_int_combined <- bind_rows(comp_int_closed, comp_int_open, comp_int_opacity) %>%
  mutate(Cultural_Variable = factor(Cultural_Variable, levels = c("Closed-Form Matches", "Open-Ended Matches", "Network Opacity (Unknowns)"), ordered = TRUE),
         close_factor = factor(close_factor, levels = rev(c("Not Close", "Somewhat Close", "Close")), ordered = TRUE))

p_int <- ggplot(comp_int_combined, aes(y = close_factor, x = estimate, fill = Cultural_Variable, color = Cultural_Variable)) +
  geom_col(width = 0.4, alpha = 0.8) +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.15, linewidth = 1) +
  scale_x_continuous(labels = scales::percent_format()) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  facet_wrap(~Cultural_Variable) +
  labs(y = "Subjective Closeness", x = "Average Marginal Effect\n(Prob. of Persistence)") +
  scale_fill_manual(values = c("Closed-Form Matches" = "#1f78b4", "Open-Ended Matches" = "#33a02c", "Network Opacity (Unknowns)" = "#e31a1c")) +
  scale_color_manual(values = c("Closed-Form Matches" = "#12476b", "Open-Ended Matches" = "#1e5e1a", "Network Opacity (Unknowns)" = "#941113")) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))

ggsave(here::here("Plots", "interaction_closeness.png"), plot = p_int, width = 12, height = 4.5)
p_int
```
As the marginal effects plot demonstrates, the moderating role of subjective closeness varies by the type of cultural measure. For closed-form matches and network opacity, there is a strong interaction with tie strength: the marginal effects are steepest for "Not Close" ties, indicating that broad cultural alignment and mutual cultural knowledge provide critical scaffolding that helps weak ties survive. Conversely, for ties that are already "Close," the slopes for closed-form matches and opacity are much flatter, suggesting these strong ties are more resilient to decay regardless of these factors. In contrast, open-ended matches do not show as strong of an interaction; their protective effect remains relatively consistent across different levels of subjective closeness.
\newpage

## Appendix: Descriptive Statistics
```{r}
#| label: tbl-descriptives
#| tbl-cap: "Descriptive Statistics"
library(modelsummary)
datasummary(
  (`Closed-Form Matches` = num_match_closed) +
  (`Open-Ended Matches` = open_match_count) + 
  (`Cultural Network Opacity` = num_unknown) + 
  (`Tie Duration (Years)` = duration_) ~ Mean + SD + Min + Max,
  data = df_period,
  output = here::here("Tabs", "desc_cont.tex")
)

datasummary(
  (`Tie Persisted` = factor(persisted)) +
  (`Ego Gender` = female_factor) + 
  (`Alter Gender` = alterfemale_factor) +
  (`Subjective Closeness` = close_factor) +
  (`Interaction Frequency` = factor(freq_factor)) +
  (`Race Homophily` = race_homophily) ~ N + Percent(),
  data = df_period,
  output = here::here("Tabs", "desc_cat.tex")
)
```


## Robustness Check: Ego Fixed Effects

To ensure the robustness of our main findings, we estimated an **Ego Fixed-Effects** model using conditional logistic regression (`survival::clogit`). This stringent test completely controls for all unobserved, time-invariant ego characteristics by only leveraging within-ego variance (i.e., comparing alters retained vs. dropped by the same ego).

```{r}
#| label: tbl-robustness-models
#| tbl-cap: "Odds Ratios for Tie Persistence (Ego Fixed-Effects Model)"
library(survival)

# Ego Fixed-Effects (Conditional Logit)
mod_fe <- clogit(
  persisted ~ num_match_closed + open_match_count + num_unknown + same_dorm + is_friend + 
    race_homophily + freq_high + alterfemale_factor + close_factor + duration_ + duration_sq + period + strata(egoid),
  data = df_period
)

modelsummary(
  list("Ego FE (clogit)" = mod_fe),
  stars = TRUE,
  exponentiate = TRUE,
  output = here::here("Tabs", "robustness_models.tex"),
  coef_map = c(
    "num_match_closed" = "Closed-Form Matches",
    "open_match_count" = "Open-Ended Matches",
    "num_unknown" = "Network Opacity",
    "close_factorSomewhat Close" = "Subjective Closeness: Somewhat",
    "close_factorClose" = "Subjective Closeness: Close",
    "same_dorm" = "Roommate/Dormmate",
    "is_friend" = "Friend",
    "freq_high" = "Frequency: Weekly/Daily",
    "race_homophilyHomophilous" = "Race Homophily: Same Race",
    "duration_" = "Tie Duration (Years)"
  ),
  gof_map = c("nobs", "aic", "bic"),
  notes = "Note: Controls for ego and alter gender, non-linear tie duration, and period fixed effects."
)
```
