---
title: "Cultural Matching Analysis"
author: "Omar Lizardo et al."
date: today
bibliography: manuscript_citations.bib
format: 
  html:
    toc: true
    keep-tex: true
    theme: cosmo
  pdf:
    toc: true
    keep-tex: true
execute:
  echo: false
  warning: false
  cache: true
---
## Overview
This analysis file examines the role of cultural matching and network opacity in predicting the persistence of social ties over time. We pool data across adjacent waves from the NetSense study to construct a person-period dataset, restricting the sample to non-kin, on-campus student ties.
We estimate the probability of tie persistence as a function of cultural matching (closed-form domains and open-ended activities) and network opacity (the number of "Don't Know" responses regarding an alter's preferences) using mixed-effects logistic regression (`lme4::glmer`).
```{r}
#| label: setup
#| include: false
library(tidyverse)
library(lme4)
library(modelsummary)
library(here)
library(broom.mixed)
ego_race <- readRDS(here("data", "processed", "ego_race.rds")) %>% mutate(sender = as.character(sender))
alter_race <- readRDS(here("data", "processed", "alter_race.rds")) %>% mutate(receiver = as.character(receiver))
df_period <- readRDS(here("data", "processed", "adjacent_waves.rds")) %>%
  filter(kin == 0, campustie == 1, !is.na(alterid)) %>%
  left_join(ego_race, by = c("egoid" = "sender")) %>%
  left_join(alter_race, by = c("alterid" = "receiver")) %>%
  mutate(
    period = factor(period),
    persisted = as.numeric(tie_persist > 0),
    female_factor = factor(female, levels = c(0, 1), labels = c("Men", "Women")),
    alterfemale_factor = factor(alterfemale, levels = c(0, 1), labels = c("Men", "Women")),
    close_factor = factor(close, levels = c(3, 2, 1), labels = c("Not Close", "Somewhat Close", "Close")),
    same_dorm = as.numeric(reltype == 3 | reltype == 4),
    is_friend = as.numeric(reltype == 6),
    race_homophily = case_when(
      is.na(ego_race) | is.na(alter_race) ~ "Unknown",
      ego_race == alter_race ~ "Homophilous",
      TRUE ~ "Heterophilous"
    ),
    race_homophily = factor(race_homophily, levels = c("Heterophilous", "Homophilous", "Unknown")),
    freq_high = if_else(freq_factor %in% c("Weekly", "Daily"), 1, 0, missing = NA_real_),
    duration_sq = duration_^2
  ) %>%
  filter(!is.na(same_dorm), !is.na(freq_high))
glmer_ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5))
ctrl_vars <- "same_dorm + is_friend + race_homophily + freq_high + female_factor * alterfemale_factor + close_factor + duration_ + duration_sq + period"
mod_closed <- glmer(
  as.formula(paste("persisted ~ num_match_closed +", ctrl_vars, "+ (1 | egoid)")),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)
mod_open <- glmer(
  as.formula(paste("persisted ~ num_match_closed + open_match_count +", ctrl_vars, "+ (1 | egoid)")),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)
mod_opac_base <- glmer(
  as.formula(paste("persisted ~ open_match_count + num_unknown +", ctrl_vars, "+ (1 | egoid)")),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)
```
## Main Effects Models
The following models assess the main effects of closed-form cultural matching, open-ended cultural matching, and network opacity on the odds of tie persistence. Model 1 shows that closed-form matches independently predict persistence. Model 2 introduces open-ended matches, demonstrating the value of incorporating both open and closed-form items. Model 3 incorporates network opacity (the number of "Don't know" responses regarding an alter's preferences). The main effects model reveals that a lack of shared cultural knowledge signals a weaker tie.
```{r}
#| label: tbl-models
#| tbl-cap: "Odds Ratios for Tie Persistence (Main Effects Models)"
modelsummary(
  list("Model 1: Closed" = mod_closed, "Model 2: + Open" = mod_open, "Model 3: + Opacity" = mod_opac_base),
  stars = TRUE,
  exponentiate = TRUE,
  output = here::here("Tabs", "main_models.tex"),
  coef_map = c(
    "num_match_closed" = "Closed-Form Matches",
    "open_match_count" = "Open-Ended Matches",
    "num_unknown" = "Network Opacity",
    "close_factorSomewhat Close" = "Subjective Closeness: Somewhat",
    "close_factorClose" = "Subjective Closeness: Close",
    "duration_" = "Tie Duration (Years)",
    "duration_sq" = "Tie Duration Squared",
    "female_factorWomen" = "Ego: Woman",
    "alterfemale_factorWomen" = "Alter: Woman",
    "female_factorWomen:alterfemale_factorWomen" = "Ego Woman × Alter Woman",
    "same_dorm" = "Roommate/Dormmate",
    "is_friend" = "Friend",
    "freq_high" = "Frequency: Weekly/Daily",
    "race_homophilyHomophilous" = "Race Homophily: Same Race",
    "race_homophilyUnknown" = "Race Homophily: Unknown",
    "periodWave 2 to 3" = "Time Period: Wave 2-3",
    "periodWave 3 to 4" = "Time Period: Wave 3-4",
    "periodWave 4 to 5" = "Time Period: Wave 4-5",
    "periodWave 5 to 6" = "Time Period: Wave 5-6",
    "periodWave 6 to 7" = "Time Period: Wave 6-7",
    "periodWave 7 to 8" = "Time Period: Wave 7-8"
  ),
  gof_map = c("nobs", "aic", "bic")
)
```
### Predicted Probabilities
**Figure @fig-combined** visualizes the predicted probability of tie persistence across the observed range of closed-form matches, open-ended matches, and network opacity.
```{r}
#| label: fig-combined
#| fig-cap: "Predicted Probability of Tie Persistence by Cultural Matches and Network Opacity"
library(ggeffects)
library(patchwork)

pred_main_closed <- ggpredict(mod_open, terms = "num_match_closed") %>% 
  as_tibble() %>% 
  mutate(Predictor_Value = x, Predictor = "Closed-Form Matches")

pred_main_open <- ggpredict(mod_open, terms = "open_match_count") %>% 
  as_tibble() %>% 
  mutate(Predictor_Value = x, Predictor = "Open-Ended Matches")

pred_main_opacity <- ggpredict(mod_opac_base, terms = "num_unknown") %>% 
  as_tibble() %>% 
  mutate(Predictor_Value = x, Predictor = "Network Opacity (Unknowns)")

pred_main_combined <- bind_rows(pred_main_closed, pred_main_open, pred_main_opacity) %>%
  mutate(Predictor = factor(Predictor, levels = c("Closed-Form Matches", "Open-Ended Matches", "Network Opacity (Unknowns)"), ordered = TRUE))

p <- ggplot(pred_main_combined, aes(x = Predictor_Value, y = predicted, color = Predictor, fill = Predictor)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  facet_wrap(~Predictor, scales = "free_x") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 0.4)) +
  scale_color_manual(values = c("Closed-Form Matches" = "#1f78b4", "Open-Ended Matches" = "#33a02c", "Network Opacity (Unknowns)" = "#e31a1c")) +
  scale_fill_manual(values = c("Closed-Form Matches" = "#1f78b4", "Open-Ended Matches" = "#33a02c", "Network Opacity (Unknowns)" = "#e31a1c")) +
  labs(y = "Predicted Probability of Tie Persistence", x = "Count of Cultural Domains / Activities") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))

ggsave(here::here("Plots", "main_effects.png"), plot = p, width = 10, height = 4.5)
p
```
## Subjective Closeness Interactions
To test whether the protective effects of cultural matching or the detrimental effects of network opacity depend on the strength of the tie, we estimated auxiliary interaction models between subjective closeness and each of the three cultural variables. While interaction terms in logistic regression models can be difficult to interpret directly from coefficients due to the non-linear link function, computing predicted probabilities across the range of the variables provides a clearer picture of the underlying dynamics.
```{r}
#| label: setup-closeness-interactions
#| include: false
ctrl_vars_int <- "same_dorm + is_friend + race_homophily + freq_high + female_factor * alterfemale_factor + duration_ + duration_sq + period"

mod_close_int_closed <- glmer(
  as.formula(paste("persisted ~ num_match_closed * close_factor +", ctrl_vars_int, "+ (1 | egoid)")),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)

mod_close_int_open <- glmer(
  as.formula(paste("persisted ~ open_match_count * close_factor +", ctrl_vars_int, "+ (1 | egoid)")),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)

mod_close_int_unknown <- glmer(
  as.formula(paste("persisted ~ num_unknown * close_factor +", ctrl_vars_int, "+ (1 | egoid)")),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)
```
**Figure @fig-closeness-interactions** plots the predicted probability of tie persistence across the range of the three cultural variables for the three levels of subjective closeness.
```{r}
#| label: fig-closeness-interactions
#| fig-cap: "Predicted Probability of Tie Persistence by Cultural Matching, Network Opacity, and Subjective Closeness"
#| fig-width: 12
#| fig-height: 5
#| warning: false
library(patchwork)
library(ggeffects)

pred_closed <- ggpredict(mod_close_int_closed, terms = c("num_match_closed", "close_factor")) %>% 
  as_tibble() %>% 
  mutate(Predictor_Value = x, close_factor = group, Cultural_Variable = "Closed-Form Matches")

pred_open <- ggpredict(mod_close_int_open, terms = c("open_match_count", "close_factor")) %>% 
  as_tibble() %>% 
  mutate(Predictor_Value = x, close_factor = group, Cultural_Variable = "Open-Ended Matches")

pred_opacity <- ggpredict(mod_close_int_unknown, terms = c("num_unknown", "close_factor")) %>% 
  as_tibble() %>% 
  mutate(Predictor_Value = x, close_factor = group, Cultural_Variable = "Network Opacity (Unknowns)")

pred_combined <- bind_rows(pred_closed, pred_open, pred_opacity) %>%
  mutate(Cultural_Variable = factor(Cultural_Variable, levels = c("Closed-Form Matches", "Open-Ended Matches", "Network Opacity (Unknowns)"), ordered = TRUE),
         close_factor = factor(close_factor, levels = c("Not Close", "Somewhat Close", "Close"), ordered = TRUE))

p_int <- ggplot(pred_combined, aes(x = Predictor_Value, y = predicted, color = close_factor, fill = close_factor)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  facet_wrap(~Cultural_Variable, scales = "free_x") +
  labs(y = "Predicted Probability of Tie Persistence", x = "Count of Cultural Domains / Activities", color = "Subjective Closeness", fill = "Subjective Closeness") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_color_viridis_d(option = "plasma", end = 0.8) +
  scale_fill_viridis_d(option = "plasma", end = 0.8) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))

ggsave(here::here("Plots", "interaction_closeness.png"), plot = p_int, width = 12, height = 5)
p_int
```
As the predicted probability plot clearly demonstrates, the role of cultural matching varies substantially by subjective closeness. For "Close" ties (the steepest slopes), an increase in shared cultural tastes—whether measured through closed-form domains or open-ended activities—dramatically increases the probability of tie persistence. Conversely, for ties that are "Not Close," the slope is essentially flat, and their baseline probability of persistence remains low regardless of matching. This suggests that shared cultural tastes provide the substantive core that allows strong ties to endure, whereas weak ties are too fleeting to be rescued by matching alone. Similarly, network opacity exerts its most severe negative effect on ties that are subjectively close.
\newpage

## Appendix: Descriptive Statistics
```{r}
#| label: tbl-descriptives
#| tbl-cap: "Descriptive Statistics"
library(modelsummary)
datasummary(
  (`Closed-Form Matches` = num_match_closed) +
  (`Open-Ended Matches` = open_match_count) + 
  (`Cultural Network Opacity` = num_unknown) + 
  (`Tie Duration (Years)` = duration_) ~ Mean + SD + Min + Max,
  data = df_period,
  output = here::here("Tabs", "desc_cont.tex")
)

datasummary(
  (`Tie Persisted` = factor(persisted)) +
  (`Ego Gender` = female_factor) + 
  (`Alter Gender` = alterfemale_factor) +
  (`Subjective Closeness` = close_factor) +
  (`Interaction Frequency` = factor(freq_factor)) +
  (`Race Homophily` = race_homophily) ~ N + Percent(),
  data = df_period,
  output = here::here("Tabs", "desc_cat.tex")
)
```


## Robustness Check: Ego Fixed Effects

To ensure the robustness of our main findings, we estimated an **Ego Fixed-Effects** model using conditional logistic regression (`survival::clogit`). This stringent test completely controls for all unobserved, time-invariant ego characteristics by only leveraging within-ego variance (i.e., comparing alters retained vs. dropped by the same ego).

```{r}
#| label: tbl-robustness-models
#| tbl-cap: "Odds Ratios for Tie Persistence (Ego Fixed-Effects Model)"
library(survival)

# Ego Fixed-Effects (Conditional Logit)
mod_fe <- clogit(
  persisted ~ num_match_closed + open_match_count + num_unknown + same_dorm + is_friend + 
    race_homophily + freq_high + alterfemale_factor + close_factor + duration_ + duration_sq + period + strata(egoid),
  data = df_period
)

modelsummary(
  list("Ego FE (clogit)" = mod_fe),
  stars = TRUE,
  exponentiate = TRUE,
  output = here::here("Tabs", "robustness_models.tex"),
  coef_map = c(
    "num_match_closed" = "Closed-Form Matches",
    "open_match_count" = "Open-Ended Matches",
    "num_unknown" = "Network Opacity",
    "close_factorSomewhat Close" = "Subjective Closeness: Somewhat",
    "close_factorClose" = "Subjective Closeness: Close"
  ),
  gof_map = c("nobs", "aic", "bic")
)
```
