## Subjective Closeness Interactions

To test whether the protective effects of cultural matching or the detrimental effects of network opacity depend on the strength of the tie, we estimated auxiliary interaction models between subjective closeness and each of the three cultural variables. While interaction terms in logistic regression models can be difficult to interpret directly from coefficients due to the non-linear link function, computing predicted probabilities across the range of the variables provides a clearer picture of the underlying dynamics.

```{r}
#| label: setup-closeness-interactions
#| include: false

mod_close_int_closed <- glmer(
  persisted ~ num_match_closed * close_factor + open_match_count + num_unknown + same_dorm + is_friend + race_homophily + freq_factor + 
    female_factor * alterfemale_factor + duration_ + period + (1 | egoid),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)

mod_close_int_open <- glmer(
  persisted ~ num_match_closed + open_match_count * close_factor + num_unknown + same_dorm + is_friend + race_homophily + freq_factor + 
    female_factor * alterfemale_factor + duration_ + period + (1 | egoid),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)

mod_close_int_unknown <- glmer(
  persisted ~ num_match_closed + open_match_count + num_unknown * close_factor + same_dorm + is_friend + race_homophily + freq_factor + 
    female_factor * alterfemale_factor + duration_ + period + (1 | egoid),
  data = df_period, family = binomial(link = "logit"), control = glmer_ctrl, nAGQ = 0
)
```

**Figure @fig-closeness-interactions** plots the marginal effects of the three cultural variables across the three levels of subjective closeness.

```{r}
#| label: fig-closeness-interactions
#| fig-cap: "Predicted Probability of Tie Persistence by Cultural Matching, Network Opacity, and Subjective Closeness"
#| fig-width: 12
#| fig-height: 5
#| warning: false

library(patchwork)

p1 <- plot_predictions(mod_close_int_closed, condition = c("num_match_closed", "close_factor"), re.form = NA) +
  labs(title = "A. Closed-Form Matches", x = "Closed-Form Matches", y = "Pr(Persistence)", color = "Subjective Closeness", fill = "Subjective Closeness") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format())

p2 <- plot_predictions(mod_close_int_open, condition = c("open_match_count", "close_factor"), re.form = NA) +
  labs(title = "B. Open-Ended Matches", x = "Open-Ended Matches", y = "Pr(Persistence)", color = "Subjective Closeness", fill = "Subjective Closeness") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format())

p3 <- plot_predictions(mod_close_int_unknown, condition = c("num_unknown", "close_factor"), re.form = NA) +
  labs(title = "C. Cultural Network Opacity", x = "Cultural Network Opacity", y = "Pr(Persistence)", color = "Subjective Closeness", fill = "Subjective Closeness") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format())

combined_close <- (p1 | p2 | p3) + plot_layout(guides = "collect") & theme(legend.position = "bottom")

ggsave(here::here("Plots", "interaction_closeness.png"), plot = combined_close, width = 12, height = 4.5)
combined_close
```

As the marginal effects plot clearly demonstrates, the role of cultural matching varies substantially by subjective closeness. For "Not Close" ties (the steepest slopes), an increase in shared cultural tastes—whether measured through closed-form domains or open-ended activities—dramatically increases the probability of tie persistence. Conversely, for ties that are already "Close," the slope is essentially flat. This suggests that shared cultural tastes provide critical scaffolding that helps weak ties survive, whereas strong ties are resilient to decay regardless of cultural matching. Similarly, cultural network opacity exerts its most severe negative effect on ties that are not subjectively close.
